opXRD: Open Experimental Powder X-ray Diffraction Database

### 2025-04-29 SI.tex
Finalized changes @ 14:41 29.04.25

\section{Description of opXRD files on Zenodo}
\replaced{The opXRD database is packaged into a single zip file ``opxrd.zip'' available for download on Zenodo.}{The database comes in two zip archives, ``opxrd.zip'' and ``opxrd\_in\_situ.zip''. The latter contains the in-situ data with highly correlated patterns recorded through time series measurements.} Within the .zip archive patterns are saved as .json files grouped in folders indicating the contributing institution. If an institution contributed data from several projects, the contributed data is further divided into folders indicating the research project. These research project folders are labeled alphabetically in the order they are introduced in Section~$3$. \added{Patterns belonging to time series measurements are labeled with filenames that indicate the measurement series they belong to and their order in that series. }

Each .json file contains a pattern recorded from an X-ray diffraction experiment\added{, as well as any available structural information of the underlying sample}. \added{At its topmost layer the .json file contains 4 mandatory fields: ``two\_theta\_values'', ``intensities'', ``powder\_experiment'' and ``metadata''. The first two together define the content of the diffraction pattern. The powder experiment field contains all available details about the experiment, including the wavelength of the employed X-ray source and everything known about the sample underlying the experiment. Insofar as these properties are available, stored sample properties include the chemical composition, lattice parameters, space group, atomic coordinates, and phase fraction of each phase. Finally, the metadata field contains information about the origin of the data including contributing institution, the name of the contributor as well as the original file format and measurement date if available.}

@@

\section{opXRD Python library \deleted{usage}}

\added{Next to the dataset on Zenodo we also offer a dedicated Python library interface with the dataset directly through Python.} The opXRD Python library allows the dataset to be accessed through one simple command: \pyth{OpXRD.load(root_dirpath)}. If the database is locally available under \pyth{root_dirpath} this command loads the library from this location. If the database is not available locally at this location, the database is automatically downloaded from Zenodo to \pyth{root_dirpath}. \added{The opXRD library also contains the code used to generate the tables and figures of this work: \url{https://github.com/aimat-lab/opXRD/blob/main/opxrd/plots.ipynb}.}

>>>>

\section{Description of opXRD files on Zenodo}
\replaced{The opXRD database is packaged as a single zip file ``opxrd.zip'' available for download on Zenodo.}{The database comes in two zip archives, ``opxrd.zip'' and ``opxrd\_in\_situ.zip''. The latter contains the in-situ data with highly correlated patterns recorded through time series measurements.} Within the .zip archive patterns are saved as .json files grouped in folders indicating the contributing institution. If an institution contributed data from several projects, the contributed data is further divided into folders indicating the research project. These research project folders are labeled alphabetically in the order they are introduced in Section~$3$. \added{Patterns belonging to time series measurements are labeled with filenames that indicate the measurement series they belong to and their order in that series. }

Each .json file contains a pattern recorded from an X-ray diffraction experiment\added{, as well as any available structural information of the underlying sample}. \added{At its topmost layer, the .json file contains 4 mandatory fields: ``two\_theta\_values'', ``intensities'', ``powder\_experiment'', and ``metadata''. The first two define the content of the diffraction pattern. The ``powder\_experiment'' field contains all available details about the experiment, including the wavelength of the employed X-ray source and everything known about the sample underlying the experiment. Insofar as these properties are available, stored sample properties include the chemical composition, lattice parameters, space group, atomic coordinates, and the phase fraction of each phase of the material. Finally, the metadata field contains information about the origin of the data, including the contributing institution, the name of the contributor as well as the original file format and measurement date if available. The metadata field will in the future also display the version of the opXRD database in which the pattern was added.}

@@

\section{Figure source code\deleted{opXRD Python library usage}}

\deleted{Next to the dataset on Zenodo we also offer a dedicated Python library interface with the dataset directly through Python. The opXRD Python library allows the dataset to be accessed through one simple command: \pyth{OpXRD.load(root_dirpath)}. If the database is locally available under \pyth{root_dirpath} this command loads the library from this location. If the database is not available locally at this location, the database is automatically downloaded from Zenodo to \pyth{root_dirpath}.}
\added{The opXRD library also contains the source code that was used to generate the tables and figures seen in Section$~3$ and the appendix. All figures, including the associated code to generate them, are located in the following notebook: \url{https://github.com/aimat-lab/opXRD/blob/9a6050ea8b62c9d67f72a126a33bf18f70cebefb/opxrd/plots.ipynb}.}

### 2025-04-29 chapters/01_introduction.tex
Finalized changes @ 14:41 29.04.25

In this work, we introduce an open powder X-ray diffraction (opXRD) database featuring a broad range of patterns collected from experiments. \added{The objective of this work is to introduce and disseminate a large, open experimental pXRD dataset, paving the way for future studies that will evaluate and benchmark its impact on model performance.} With a total of $92,552$ patterns collected from 6 contributing institutions, the opXRD database exceeds the size of the previously largest database of openly accessible experimental powder diffraction data by two orders of magnitude. To the best of our knowledge, the largest database of this type is the RRUFF database, containing 1290 experimental powder diffraction patterns \cite{lafuente2015}. Larger commerical datasets such as the PDF5+\cite{GatesRector2019} and the Linus Pauling File\cite{villars2018} exist, but their utility is limited by fees and restrictive licenses. License terms of commercial datasets, such as the PDF5+ and the Linus Pauling File, prohibit or restrict the publication of models trained on their data. In contrast, the opXRD database is both free and imposes no restrictions on how its data is used. Fig.~$\eqref{fig:ml_uses}$ provides an overview of machine learning workflows enabled and supported by the opXRD database.

@@

Of the \numpatterns patterns in the opXRD database\added{,} \numlabeled patterns come with at least partial structural information of the underlying sample. Of these \numlabeled labeled patterns, more than 900 have full structural labels including atomic coordinates. \deleted{This constitutes an experimental pXRD test dataset that is larger in size, richer in labels, and broader in represented experimental setups than the RRUFF database, which only provides lattice parameters as labels\cite{Armbruster2015}.} 
\added{While the fraction of samples with structural labels (\SI{2.35}{\percent}) may appear small, this fraction represents the largest openly available collection of experimentally derived, structure-annotated pXRD patterns. As a comparison, the RRUFF dataset, often used for benchmarking ML models, contains partial labels for \num{1290} patterns, but no atomic coordinates. The pXRD dataset is larger in size, richer in labels, and broader in represented experimental setups compared to previous openly available datasets. Given the inherently labor-intensive nature of manual labeling in pXRD analysis, it is impractical to expect a fully labeled dataset at the scale of simulated training datasets (exceeding $10^5$ patterns). Therefore, next to the option of benchmarking models and methods on the labeled fraction, opXRD is designed to complement vast simulated datasets. This can be achieved through the adjustment of simulation parameters by comparing with the unlabeled fraction of the dataset, and through transfer learning strategies.
We now want to discuss these two options of utilizing the unlabeled portion of our dataset in more detail.}
\deleted{However, since the majority of the opXRD database is unlabeled we also want to further discuss the uses of unlabeled data, including its role in improving pattern simulations and its application in transfer learning approaches.}

>>>>

In this work, we introduce an open powder X-ray diffraction (opXRD) database featuring a broad range of patterns collected from experiments.\added{The objective of this work is to introduce and disseminate a large, open experimental pXRD dataset, paving the way for future studies that will evaluate and benchmark its impact on model performance.} With a total of $\numpatterns$ patterns collected from 6 contributing institutions, the opXRD database exceeds the size of the previously largest database of openly accessible experimental powder diffraction data by two orders of magnitude. To the best of our knowledge, the largest database of this type is the RRUFF database, containing 1290 experimental powder diffraction patterns \cite{lafuente2015}. Larger commerical datasets such as the PDF5+\cite{GatesRector2019} and the Linus Pauling File\cite{villars2018} exist, but their utility is limited by fees and restrictive licenses. License terms of commercial datasets, such as the PDF5+ and the Linus Pauling File, prohibit or restrict the publication of models trained on their data. In contrast, the opXRD database is both free and imposes no restrictions on how its data is used. Fig.~$\eqref{fig:ml_uses}$ provides an overview of machine learning workflows enabled and supported by the opXRD database.

@@

Of the $\numpatterns$ patterns in the opXRD database\added{,} \numlabeled patterns come with at least partial structural information of the underlying sample. Of these \numlabeled labeled patterns, \numfull have full structural labels including lattice and basis. \deleted{This constitutes an experimental pXRD test dataset that is larger in size, richer in labels, and broader in represented experimental setups than the RRUFF database, which only provides lattice parameters as labels\cite{Armbruster2015}.} 
\added{While the fraction of samples with structural labels (\SI{2.35}{\percent}) may appear small, this fraction represents the largest openly available collection of experimentally derived, structure-annotated pXRD patterns. As a comparison, the RRUFF database, often used for benchmarking ML models, contains partial labels for \num{1290} patterns, but no atomic coordinates. The pXRD dataset is larger in size, richer in labels, and broader in represented experimental setups compared to previous openly available datasets. Given the inherently labor-intensive nature of manual labeling in pXRD analysis, it is impractical to expect a fully labeled dataset at the scale of simulated training datasets, which commonly exceed $10^6$ patterns \cite{Salgado2023, Schopmans2023}. Therefore, next to the option of benchmarking models and methods on the labeled subset, opXRD is designed to complement vast simulated datasets. This can be achieved through the adjustment of simulation parameters by comparing with the unlabeled fraction of the dataset, and through transfer learning strategies.
We now want to discuss these two options of utilizing the unlabeled portion of our dataset in more detail.}
\deleted{However, since the majority of the opXRD database is unlabeled we also want to further discuss the uses of unlabeled data, including its role in improving pattern simulations and its application in transfer learning approaches.}

### 2025-04-29 chapters/04_our_dataset.tex
Finalized changes @ 14:41 29.04.25

In collaboration with several other research institutions, we have collected a database of $\numpatterns$ experimental patterns of which 2179 are at least partially labeled with structural information of the underlying material. The following research institutions contributed data to the opXRD database: The French National Centre for Scientific Research (CNRS), Hong Kong University of Science and Technology (Guangzhou) (HKUST), University of Southern California (USC), Lawrence Berkeley National Laboratory (LBNL), Empa–Swiss Federal Laboratories for Materials Science and Technology (EMPA) and the Karlsruhe Institute of Technology (KIT). Tab.~$\eqref{tab:merged}$ provides an overview of the contributions of each institution. We filtered the submitted datasets to exclude patterns with invalid features such as only one unique recorded angle, negative angles, less than 50 recorded angles total, or all intensities being zero. \added{We standardized the associated structural information according to the standards described by Setyawan {\it et al.}\cite{Setyawan2010} using \textsc{pymatgen}. In particular, this standardization enforces a single crystal axis convention throughout all opXRD data. }

\begin{table}[!htb]
\centering
\caption{\footnotesize Overview of the contributions to the opXRD database: The availability of the chemical composition, spacegroups, lattice parameters, and atomic coordinates of the underlying samples are indicated by the columns ``Comp.'', ``Spg.'', ``Lattice'' and ``\replaced{Basis}{ Atom coords}.'' respectively.}
\label{tab:merged}
\scalebox{0.82}{
\begin{adjustbox}{center}
\begin{NiceTabular}{@{}llllllll@{}}[columns-width=1.2cm]
%\begin{tabular}{@{}llccccl@{}}
\toprule
\textbf{\ Institution} & \textbf{No. patterns} & \textbf{Comp.} & \textbf{Spg.} & \textbf{Lattice} & \textbf{Basis} & \textbf{Research Project} \\
\midrule
\ CNRS        &  1052         &  \cmark   & \pchk{85}  & \cmark    &   \pchk{85}   & Diffraction data extracted from the COD \\ 
\ USC         &  338          &  \cmark   & \cmark     & \pchk{90} &   \xmark   & Study of CuNi and CuAl alloys          \\
\ HKUST(GZ)   &  520          &  \pchk{4} & \pchk{4}   & \pchk{4}  &   \pchk{4}   & Phase identification dataset  \\ 
\ EMPA        &  770          & \cmark    & \pchk{63}  & \xmark    &   \xmark   & Metal halide perovskites, Zn-V-N libraries \\ 
\ INT         &  19,796       & \xmark    & \xmark     & \xmark    &   \xmark   & Compilation of various projects        \\ 
\ IKFT        &  64           & \xmark    & \xmark     & \xmark    &   \xmark   & Commercial catalysts, metals, metal oxides   \\ 
\ LBNL        &  70,012       & \xmark    & \xmark     & \xmark    &   \xmark   & Perovskites precursors, Mn-Sb-O system  \\ 
\midrule
\midrule
\added{\ $\sum$ Labeled}   & 2,179  & \pchk{79}  & \pchk{66}  & \pchk{63}  & \pchk{42} & Partially and fully labeled opXRD data \\
\added{\ $\sum$ Unlabeled }& 90,373 & \xmark     & \xmark     & \xmark     & \xmark    & Unlabeled opXRD data\\
\bottomrule
\end{NiceTabular}
\end{adjustbox}}
\end{table}

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=\linewidth]{figures/hist_XRD.png}
    \caption{Histograms detailing properties of all diffraction patterns in the opXRD database: a) distribution of angular resolutions; b) distribution of smallest and largest recorded $2\theta$ values.}
    \label{fig:hist_XRD}
\end{figure}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=\linewidth]{figures/hist_CRYSTAL.png}
    \caption{Histograms detailing properties of the structures underlying labeled diffraction patterns in the opXRD database: a) distribution of structures containing specified elements;b) distribution of structures containing spacegroups; c) distribution of No. atoms $N_{\text{atom}}$ contained in unit cell; d) distribution of unit cell volume $V_{\text{uc}}$}
    \label{fig:hist_CRYSTAL}
\end{figure}

>>>>

In collaboration with several other research institutions, we have collected a database of $\numpatterns$ experimental \added{powder diffraction} patterns. \added{Of these patterns, $\numlabeled$} are at least partially labeled with structural information \added{and $\numfull$ are fully labeled through the lattice and basis of the underlying material}. The following research institutions contributed data to the opXRD database: The French National Centre for Scientific Research (CNRS), Hong Kong University of Science and Technology (Guangzhou) (HKUST), University of Southern California (USC), Lawrence Berkeley National Laboratory (LBNL), Empa–Swiss Federal Laboratories for Materials Science and Technology (EMPA) and the Karlsruhe Institute of Technology (KIT). \added{We have taken measures to ensure data validation through both manual and automated processes. Before parsing the data we established the file formats and data organization of each submission to ensure that the files were compatible with our custom parsing mechanism. As part of the automatic parsing process,} we filtered the submitted datasets to exclude patterns with invalid features such as only one unique recorded angle, negative angles, less than 50 recorded angles total, or all intensities being zero. \added{After the parsing, we also manually inspected a random selection of patterns from each submitted dataset for any anomalies that would warrant further investigation.}

\added{We standardized the associated structural information according to the standards described by Setyawan {\it et al.}\cite{Setyawan2010} using \textsc{pymatgen}. In particular, this standardization enforces a single crystal axis convention throughout all opXRD data. } Tab.~$\eqref{tab:merged}$ provides an overview of the contributions of each institution.

\begin{table}[!htb]
\centering
\caption{\footnotesize Overview of the contributions to the opXRD database: The availability of the chemical composition, spacegroups, lattice parameters, and atomic coordinates of the underlying samples are indicated by the columns ``Comp.'', ``Spg.'', ``Lattice'' and ``\replaced{Basis}{ Atom coords}.'' respectively.}
\label{tab:merged}
\scalebox{0.81}{
\begin{adjustbox}{center}
\begin{NiceTabular}{@{}llllllll@{}}[columns-width=1.2cm]
\toprule
\textbf{\ Institution} & \textbf{No. patterns} & \textbf{Comp.} & \textbf{Spg.} & \textbf{Lattice} & \textbf{Basis} & \textbf{Research Project} \\
\midrule
\ CNRS        &  1052         &  \cmark   & \pchk{85}  & \cmark    &   \pchk{85}   & Diffraction data extracted from the COD \\ 
\ USC         &  338          &  \cmark   & \cmark     & \pchk{90} &   \xmark   & Study of CuNi and CuAl alloys          \\
\ HKUST(GZ)   &  520          &  \pchk{4} & \pchk{4}   & \pchk{4}  &   \pchk{4}   & Phase identification dataset  \\ 
\ EMPA        &  770          & \cmark    & \pchk{63}  & \xmark    &   \xmark   & Metal halide perovskites, Zn-V-N libraries \\ 
\ INT         &  19,796       & \xmark    & \xmark     & \xmark    &   \xmark   & Compilation of various projects        \\ 
\ IKFT        &  64           & \xmark    & \xmark     & \xmark    &   \xmark   & Commercial catalysts, metals, metal oxides   \\ 
\ LBNL        &  70,012       & \xmark    & \xmark     & \xmark    &   \xmark   & Perovskites precursors, Mn-Sb-O system  \\ 
\midrule
\midrule
\added{\ $\sum$ Labeled}   & \added{2,179}  & \pchk{79}  & \pchk{66}  & \pchk{63}  & \pchk{42} & \added{Partially and fully labeled opXRD data} \\
\added{\ $\sum$ Unlabeled }& \added{90,373} & \xmark     & \xmark     & \xmark     & \xmark    & \added{Unlabeled opXRD data}\\
\bottomrule
\end{NiceTabular}
\end{adjustbox}}
\end{table}

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=\linewidth]{figures/hist_XRD.png}
    \caption{\added{Histograms detailing properties of all diffraction patterns in the opXRD database: a) distribution of angular resolutions, b) distribution of smallest and largest recorded $2\theta$ values.}}
    \label{fig:hist_XRD}
\end{figure}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=\linewidth]{figures/hist_CRYSTAL.png}
    \caption{\added{Histograms detailing properties of the structures underlying labeled diffraction patterns in the opXRD database: a) distribution of structures containing specified elements, b) distribution of structures containing spacegroups, c) distribution of No. atoms $N_{\text{atom}}$ contained in unit cell, d) distribution of unit cell volume $V_{\text{uc}}$.}}
    \label{fig:hist_CRYSTAL}
\end{figure}

### 2025-04-29 chapters/05_how_to_use.tex
Finalized changes @ 14:41 29.04.25

Next to the availability of the opXRD dataset on Zenodo, we also provide a Python library ``opxrd'' to easily download and interface with the dataset. The instructions for how to install this library can be found in the repository associated with the library. The repository to this library is located at \url{https://github.com/aimat-lab/opxrd}. The opxrd library includes options for data-loading, standardization, plotting, and the conversion to \emph{PyTorch} tensors. We provide a Jupyter Notebook (\url{https://colab.research.google.com/github/aimat-lab/opXRD/blob/main/opxrd/usage.ipynb})
that showcases these functionalities in more detail. This notebook also illustrates how to interface with the opXRD database through Python.

>>>>

\replaced{We}{Next to the availability of the opXRD dataset on Zenodo, we} also provide a Python library ``opxrd'' to easily download and interface with the dataset.
\added{The opXRD library is designed for easy integration with common machine learning frameworks such as \emph{PyTorch}. This makes it an ideal resource for researchers developing and benchmarking sim-to-real transfer strategies in pXRD data analysis.}
The instructions for how to install this library can be found in the repository associated with the library. The repository to this library is located at \url{https://github.com/aimat-lab/opxrd}. The opxrd library includes options for data-loading, standardization, plotting, and the conversion to \emph{PyTorch} tensors. We provide a Jupyter Notebook (\url{https://colab.research.google.com/github/aimat-lab/opXRD/blob/main/opxrd/usage.ipynb})
that showcases these functionalities in more detail. This notebook also illustrates how to interface with the opXRD database through Python.

### 2025-04-29 chapters/06_summary_and_outlook.tex
Finalized changes @ 14:41 29.04.25

With the opXRD database, a curation of $\numpatterns$ unlabeled and 2179 at least partially labeled experimental powder X-ray diffraction patterns from a wide range of different materials systems, we provide the largest currently available source of experimental XRD patterns. With this, we address the need for experimental data that arises when developing algorithms and analysis tools for pXRD data, both based on machine learning and classical approaches. The data can be used for the actual method development and for testing. Our dataset is a valuable and so far missing resource to drive further developments in the automated analysis of XRD data.

% How to contribute
Rather than a finished project, the opXRD database is an ongoing effort to collect experimental powder XRD data. We invite everyone who is working in the area of experimental powder XRD to submit it to the dataset, in order to further improve the utility of the dataset and thus aid further developments in this field. Our submission page (\url{https://xrd.aimat.science/}) and submission helper software will be kept available to collect more data. We will keep updating and maintaining the dataset with new incoming submissions.

>>>>

With the opXRD database, a curation of $\numpatterns$ unlabeled and 2179 at least partially labeled experimental powder X-ray diffraction patterns from a wide range of different materials systems, we provide the largest currently available source of experimental XRD patterns. With this, we address the need for experimental data that arises when developing algorithms and analysis tools for pXRD data, both based on machine learning and classical approaches. The data can be used for the actual method development and for testing. Our dataset is a valuable and so far missing resource to drive further developments in the automated analysis of XRD data. \added{Looking forward, opXRD is expected to play an important role in the development of advanced transfer learning approaches that integrate large-scale simulated data with real experimental patterns, ultimately driving the automation and accuracy of pXRD analysis. Future work will also include comprehensive benchmark evaluations to quantify the performance improvements achieved by incorporating opXRD into transfer learning pipelines.}

% How to contribute
Rather than a finished project, the opXRD database is an ongoing effort to collect experimental powder XRD data. We invite everyone \deleted{who is} working in the area of experimental powder XRD to submit it to the dataset, \deleted{in order} to further improve the utility of the dataset and thus aid further developments in this field. Our submission page (\url{https://xrd.aimat.science/}) \deleted{and submission helper software} will \replaced{continue to stay}{ be kept available} \added{available for the submission of data. As new submissions come in, newer versions of the opXRD database incorporating these submitted datasets will be released.}
\added{As the opXRD database grows further, we look forward to expanding this website to become a comprehensive community resource from which the database can be governed. Planned resources on this site include a contributor list, version list, and changelogs as well as comprehensive versioning, license, citation, and attribution practice statements. }

We will keep updating and maintaining the dataset with new incoming submissions.
